---
title: "Appendix S5 Case study 3 Forest cover "
author: "Bob Douma & James Weedon"
date: "4 February 2019"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```




This appendix has two parts. The first part shows how the data that was used for Case Study 3 was made. The second part shows the relationship quadrat size and the root mean squared error of various beta regression models.


## Loading libraries
```{r, suppressWarnings=T}
library(rgdal)
library(raster)
library(sp)
library(RODBC)
library(maptools)
library(RArcInfo)
library(betareg)
library(gtools)
library(bbmle)
library(VGAM)
library(reshape2)
```

## Defining user defined functions
```{r, suppressWarnings=T}
# function to transform fraction with 0 and 1 to <1 and >0
transform01 <- function(x) {
  (x * (length(x) - 1) + 0.5) / (length(x))
}
# function to backtransform transform01
backtransform.est <- function(x, n) {
  y <- (x * n - 0.5) / (n - 1)
  return(y)
}

RMSE <- function(x, y) {
  a <- sqrt(sum((y - x)^2) / length(x))
  return(a)
}
```


## Code to generate forests

## function to generate square grid of size gridsize (meters)
```{r, suppressWarnings=T,eval=T}
forest <- function(gridsize) {
  square <- matrix(c(0, 0, 0, gridsize, gridsize, gridsize, gridsize, 0, 0, 0), 
                   ncol = 2, byrow = 2)
  square.1 <- Polygon(square, hole = FALSE)
  square.polygons <- Polygons(list(square.1), ID = c("1"))
  square <- SpatialPolygons(list(square.polygons))
  area.square <- data.frame(area = sapply(square@polygons,
    FUN = function(x) {
      slot(x, "area")
    }
  )) # calculates area of square
  return(list(square, area.square))
}
```



## generate a number of f trees in an area of the gridsize.
```{r, suppressWarnings=T,eval=T}
trees <- function(gridsize, f) {
  # generate position of trees in square; random assignment of location
  x <- runif(f, min = 0, max = gridsize)
  y <- runif(f, min = 0, max = gridsize)
  coord <- data.frame(x, y)
  #  points(coord)


  pols <- list()
  for (i in 1:f) { # loop to generate tree polygons
    R.j <- rlnorm(1, meanlog = 1, sdlog = 0.2)
    # create points on circle polygon with radius R
    xcircle <- coord[i, 1] + R.j * cos(seq(0, 2 * pi, length.out = 100))
    ycircle <- coord[i, 2] + R.j * sin(seq(0, 2 * pi, length.out = 100))
    # close circle exactly
    xcircle[100] <- xcircle[1]
    ycircle[100] <- ycircle[1]
    # make Polyon from points
    disp.distance <- Polygon(cbind(xcircle, ycircle), hole = FALSE)
    disp.distance.Polygons <- Polygons(list(disp.distance), ID = c(i))
    disp.distance.Polygons <- SpatialPolygons(list(disp.distance.Polygons))

    pols[[i]] <- disp.distance.Polygons
    pols[[i]] <- spChFIDs(pols[[i]], paste("a", i, row.names(pols[[i]]), sep = "."))
  }

  joined <- SpatialPolygons(lapply(
    pols,
    function(x) {
      x@polygons[[1]]
    }
  )) # joins all tree in a single sp object
  union <- aggregate(unionSpatialPolygons(joined,
    ID = paste("a", seq(1, f))
  )) # makes union of all trees
  return(union)
}
```

```{r, suppressWarnings=T,eval=T}
# generate and plot area
fore <- forest(100)
plot(fore[[1]])
# generate trees
trees.for <- trees(100, 80)
plot(trees.for, add = T)
```

```{r, suppressWarnings=T,eval=T}
treecover <- function(union, square, gridsize) {
  union.area <- data.frame(area = sapply(union@polygons,
    FUN = function(x) {
      slot(x, "area")
    }
  )) # make
  intersect.area <- intersect(union, square) # calculate intersection area

  intersect.area <- data.frame(area = sapply(intersect.area@polygons,
    FUN = function(x) {
      slot(x, "area")
    }
  ))
  true.cover <- intersect.area / area(square)

  return(true.cover)
}
```

## calculate tree cover
```{r , suppressWarnings=T,eval=T}
treecover(trees.for, fore[[1]], 100)
```

## sample cover in quadrats
The percent cover in the `n` quadrats is calculated based on the quadrat size and the trees in the area.
```{r,suppressWarnings=T,eval=T}
quadrats <- function(trees, quadsize, gridsize, n, overlapping = F) {
  size <- rep(quadsize, n)
  if (overlapping) {
    # assign location of quadrates randomly
    x <- runif(n, min = 0, max = 100 - size[1])
    y <- runif(n, min = 0, max = 100 - size[1])
  } else {
    # quadrats are located on a grid, but chosen randomly
    x <- seq(0, gridsize - quadsize, quadsize)
    y <- seq(0, gridsize - quadsize, quadsize)
    xy <- expand.grid(x, y)
    xy <- xy[sample(c(1:nrow(xy)), n, replace = F), ]
    x <- xy[, 1]
    y <- xy[, 2]
  }

  # loop of quadrats to determine percent cover in each quadrat
  fCover <- numeric(length = n)
  for (k in 1:n) {

    # make quadrat
    square <- matrix(c(
      x[k], y[k], x[k], (y[k] + size[k]),
      (x[k] + size[k]), (y[k] + size[k]), (x[k] + size[k]), y[k], x[k], y[k]
    ), ncol = 2, byrow = 2)
    square.1 <- Polygon(square, hole = FALSE)
    # make polygon of quadrat
    square.polygons <- Polygons(list(square.1), ID = c("1"))
    square.quadrat <- SpatialPolygons(list(square.polygons))
    # calculate intersection of quadrat and trees
    intersect.area <- intersect(trees, square.quadrat)
    if (is.null(intersect.area)) {
      fCover[k] <- 0
    } else {
      intersect.areas <- data.frame(area = sapply(intersect.area@polygons,
        FUN = function(x) {
          slot(x, "area")
        }
      ))
      area.square.r <- data.frame(area = sapply(square.quadrat@polygons,
        FUN = function(x) {
          slot(x, "area")
        }
      ))
      fCover[k] <- intersect.areas / area.square.r
    }
  }

  fCover <- unlist(fCover)
  if (length(fCover) != n) {
    print(c("number of values not equal to number of quadrats", fCover[which(fCover > 1)]))
  }
  return(fCover)
}
```

# set simulation characteristics
```{r, suppressWarnings=T, warning=F,message=F,eval=T}
n <- 15 # number of quadrats positioned in grid
quadsize <- 10 # square meter
gridsize <- 100 # length of grid (100m)
reps <- 20 # number of forests
MAP <- c(125, 2500)
# we assume 0.4 tree per mm of rainfall/year
MAP <- ceiling(seq(MAP[1], MAP[2], length.out = reps))
# number of plants sown in grid
nplants <- ceiling(seq(MAP[1], MAP[length(MAP)], length.out = reps)) * 0.4  
```

# create dataframe to store results in
```{r, suppressWarnings=T , warning=F,message=F,eval=T}
fCover <- as.data.frame(matrix(NA, nrow = reps, 5 + n - 1))
```

# simulate forests
```{r,suppressWarnings=T, warning=F,message=F,eval=T, cache=TRUE}
forests <- forest(gridsize)[[1]] # generate empty forest area

count <- 0
for (i in nplants) {
  count <- count + 1
  # generates union of trees
  trees.f <- trees(gridsize, i) 
  # calculates treecover over whole area
  treecov <- treecover(trees.f, forests, gridsize) 
  # calculate tree cover in n quadrats
  quadcov <- quadrats(trees.f, quadsize, gridsize, n, overlapping = F) 
  # store results
  fCover[count, 1] <- i * 1 / 0.4 # to get MAP
  fCover[count, 2] <- gridsize
  fCover[count, 3] <- quadsize
  fCover[count, 4] <- treecov
  fCover[count, 5:(4 + n)] <- quadcov
}
```
## Analysing results
```{r, suppressWarnings=T,eval=T}
fCover.m <- melt(fCover, id.var = 1:4)
```

## Fitting statistical model to percent cover
```{r, suppressWarnings=T,eval=T}
fCover.m$covert <- transform01(fCover.m[, 6])
bet1 <- betareg(fCover.m$covert ~ fCover.m[, 1] | fCover.m[, 1], link = "cloglog")
bet2 <- betareg(fCover.m$covert ~ fCover.m[, 1] | fCover.m[, 1], link = "logit")
bet3 <- betareg(1 - fCover.m$covert ~ fCover.m[, 1] - 1 | fCover.m[, 1], link = "log")
lm1 <- lm(logit(fCover.m$covert) ~ fCover.m[, 1])
lm2 <- lm(log(1 - fCover.m$covert) + 0 ~ fCover.m[, 1])
coef.bet1 <- coef(bet1)
coef.bet2 <- coef(bet2)
coef.bet3 <- coef(bet3)
coef.lm <- coef(lm1)
coef.lm2 <- coef(lm2)
```

## Plotting % cover as a function of number of individuals in the forest area.
```{r, suppressWarnings=T,eval=T}

plot(fCover.m[, "covert"] ~ fCover.m[, 1], xlab = "MAP (mm/year)", 
     ylab = "percent cover", col = scales::alpha("grey", 0.7))
points(fCover.m[, 4] ~ fCover.m[, 1], col = "blue", lwd = 2)
curve(backtransform.est(cloglog(coef.bet1[1] + coef.bet1[2] * x, inverse = T), 
                        nrow(fCover.m)),
  col = "orange", lwd = 2, lty = 1, add = T
)
curve(backtransform.est(inv.logit(coef.bet2[1] + coef.bet2[2] * x), nrow(fCover.m)),
  add = T,
  col = "purple", lwd = 2
)
curve(backtransform.est(1 - exp(coef.bet3[1] * x), nrow(fCover.m)),
  add = T,
  col = "green", lwd = 2, lty = 1
)
curve(backtransform.est(inv.logit(coef.lm[1] + coef.lm[2] * x), nrow(fCover)),
  add = T,
  col = "red", lwd = 2
)
curve(backtransform.est(1 - exp(coef.lm2[1] + coef.lm2[2] * x), nrow(fCover)),
  add = T,
  col = "red", lwd = 2, lty = 2
)
legend("bottomright", c("true cover", "beta-cloglog", "beta-logit", "beta-log",
                        "logit trans", "log trans"),
  col = c("blue", "orange", "purple", "green", "red", "red"),
  lty = c(NA, 1, 1, 1, 1, 2), pch = c(1, NA, NA, NA, NA, NA), lwd = 2
)
```
## Calculating RMSE of different models
```{r, suppressWarnings=T,eval=T}

fCover.m$fit.beta <- fitted(bet1)
fCover.m$fit.beta2 <- fitted(bet2)
fCover.m$fit.beta3 <- 1 - fitted(bet3)
fCover.m$logit1 <- inv.logit(fitted(lm1))
fCover.m$log2 <- exp(fitted(lm2))


rmse.beta.cloglog <- RMSE(fCover.m$V4, fCover.m$fit.beta)
rmse.beta.logit <- RMSE(fCover.m$V4, fCover.m$fit.beta2)
rmse.beta.log <- RMSE(fCover.m$V4, fCover.m$fit.beta3)
rmse.logit <- RMSE(fCover.m$V4, fCover.m$logit1)
rmse.log <- RMSE(fCover.m$V4, fCover.m$log2)

# ratio of RMSE for best fitting beta model and transformation
rmse.beta.log / rmse.logit
```

## Test if mismatch between model predictions and observations is indeed substantially smaller in case of beta regression by generating many forests.

To explore the effect of the quadrat size on the average root mean squared error we varied the quadrat size in 9 steps from 4 to 20.

```{r,eval=F}
eff.quadrat.size <- function(quadsize, sims, n = 15, nplants, ndep = F, area.sam = 1600) {
  if (area.sam > gridsize^2) {
    print(c("you want to sample a larger area than the total forest area"))
  }

  if (ndep) {
    n <- round(area.sam / (quadsize^2))
  }
  fCover <- as.data.frame(matrix(NA, nrow = length(nplants), 5 + n - 1))
  print(quadsize)
  # make data frame to store results in
  out <- as.data.frame(matrix(NA, nrow = sims, ncol = 5))
  for (j in 1:sims) {
    forests <- forest(gridsize)[[1]] # generate empty forest area

    count <- 0
    for (i in nplants) {
      count <- count + 1
      # generates union of trees
      trees.f <- trees(gridsize, i) 
      # calculates treecover over whole area
      treecov <- treecover(trees.f, forests, gridsize) 
      # calculate tree cover in n quadrats
      quadcov <- suppressWarnings(quadrats(trees.f, quadsize, 
                                           gridsize, n = n, overlapping = F))
      # store results
      fCover[count, 1] <- i
      fCover[count, 2] <- gridsize
      fCover[count, 3] <- quadsize
      fCover[count, 4] <- treecov

      if (sum(quadcov > 1)) {
        print(c("some quadrat covers larger than one, set to a maximum of one",
                quadcov[which(quadcov > 1)]))
        quadcov[quadcov > 1] <- 1
      }
      fCover[count, 5:(4 + n)] <- quadcov
    }

    if (sum(fCover, na.rm = T) == 0) {
      print("only zeros or ones")
      # store results
      out[j, 1] <- NA
      out[j, 2] <- NA
      out[j, 3] <- NA
      out[j, 4] <- NA
      out[j, 5] <- NA
      out1 <- data.frame(quadsize, j, out[j, ])
    } else {


      # melt data.frame
      fCover.m <- melt(fCover, id.var = 1:4)

      fCover.m$covert <- transform01(fCover.m[, 6])
      bet1 <- suppressWarnings(betareg(fCover.m$covert ~ fCover.m[, 1] | fCover.m[, 1], 
                                       link = "cloglog"))
      bet2 <- suppressWarnings(betareg(fCover.m$covert ~ fCover.m[, 1] | fCover.m[, 1], 
                                       link = "logit"))
      bet3 <- suppressWarnings(betareg(1 - fCover.m$covert ~ fCover.m[, 1] - 1 | fCover.m[, 1],
                                       link = "log"))
      lm1 <- lm(logit(fCover.m$covert) ~ fCover.m[, 1])
      lm2 <- lm(log(1 - fCover.m$covert) + 0 ~ fCover.m[, 1])
      coef.bet1 <- coef(bet1)
      coef.bet2 <- coef(bet2)
      coef.bet3 <- coef(bet3)
      coef.lm <- coef(lm1)
      coef.lm2 <- coef(lm2)

      # calculate fitted values
      if (length(fitted(bet1)) == nrow(fCover.m)) {
        fCover.m$fit.beta <- fitted(bet1)
      } else {
        (fCover.m$fit.beta <- NA)
      }

      if (length(fitted(bet2)) == nrow(fCover.m)) {
        fCover.m$fit.beta2 <- fitted(bet2)
      } else {
        (fCover.m$fit.beta2 <- NA)
      }

      if (length(fitted(bet3)) == nrow(fCover.m)) {
        fCover.m$fit.beta3 <- 1 - fitted(bet3)
      } else {
        (fCover.m$fit.beta3 <- NA)
      }

      fCover.m$logit1 <- inv.logit(fitted(lm1))
      fCover.m$log2 <- exp(fitted(lm2))

      # calculate RMSE. # root mean squared error of true cover and 
      # modelled tree cover of a 1 ha plot.
      rmse.beta.cloglog <- RMSE(fCover.m$V4, fCover.m$fit.beta)
      rmse.beta.logit <- RMSE(fCover.m$V4, fCover.m$fit.beta2)
      rmse.beta.log <- RMSE(fCover.m$V4, fCover.m$fit.beta3)
      rmse.logit <- RMSE(fCover.m$V4, fCover.m$logit1)
      rmse.log <- RMSE(fCover.m$V4, fCover.m$log2)
    }
    # store results
    out[j, 1] <- rmse.beta.cloglog
    out[j, 2] <- rmse.beta.logit
    out[j, 3] <- rmse.beta.log
    out[j, 4] <- rmse.logit
    out[j, 5] <- rmse.log
    out1 <- data.frame(quadsize, j, out[j, ])
  }

  ## mean bias
  # average root mean squared error of true cover and modelled tree cover of a 1 ha plot.
  mismatch <- apply(out, 2, mean)
  names(mismatch) <- c("cloglog", "logit", "log", "logit-t", "log-t")
  return(mismatch)
}

n <- 15 # number of quadrats positioned in grid
quadsize <- 10 # square meter
gridsize <- 100 # length of grid (100m)
reps <- 20 # number of forests
MAP <- c(125, 2500)
# we assume 0.4 tree per mm of rainfall/year
MAP <- ceiling(seq(MAP[1], MAP[2], length.out = reps))
# number of plants sown in grid
nplants <- ceiling(seq(MAP[1], MAP[length(MAP)], length.out = reps)) * 0.4 
# specify quadrat sizes
quads <- c(4, 6, 8, 10, 15, 20) # seq(2,50,length.out = 9)
# calculation may take long when sims > 5
rmse.quadsize <- sapply(quads, eff.quadrat.size, sims = 5, n = 15, nplants = nplants, ndep = T) 
rmse.quadsize
```


![Root mean squaed error as a function of quadrat size for beta regression with various link functions (cloglog, log, logit) and the logit transformation (logit.t). The log transformation is not shown because the RMSE is approximately a factor ten higher compared to the other models. Number of quadrats is kept constant while varying qaudrat size. See main text for details. Error bars represent standard errors of the mean, n = 50](quadsize_rmse.pdf)


![Root mean squaed error as a function of quadrat size for beta regression with various link functions  (cloglog, log, logit) and the logit transformation (logit.t). The log transformation is not shown because the RMSE is approximately a factor ten higher compared to the other models.Total  area sampled is kept constant while varying qaudrat size.  See main text for details. Error bars represent standard errors of the mean, n = 50](quadsize_areaconstant_rmse.pdf)

